Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-25 paper.qmd
Added screenshots for simulated data, videos

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
pcentroid_plot_s_curve <- ggplot(data = hex_grid, aes(x = x, y = y)) +
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = df_bin_centroids, aes(x = c_x, y = c_y), color = "#33a02c") +
  coord_equal() +
  theme_linedraw() +
  theme(legend.position = "bottom", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

@@

cell_count_plot <- ggplot(df_bin_centroids_s_curve, aes(x = reorder(as.factor(hexID), -std_counts), y = std_counts)) +
  geom_quasirandom() + xlab("hexagonal id") + ylab("Standardized cell count") +
  geom_hline(yintercept = 0.255, colour = "#de2d26") +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 7),
        axis.text.x = element_text(angle = 90))

@@

distance_plot <- plot_dist(distance_s_curve) +
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12))
# 
# distance_plot

@@

@@

trimesh_removed_s_curve_umap <- trimesh_removed_s_curve_umap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)
```

@@

```{r}
#| echo: false
#| fig-cap: "Distribution of 2D Euclidean distances."
#| label: fig-stdctsScurve

@@

<!--
```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false

hex_full_count_df_loop2 <- read_rds("data/s_curve/s_curve_hex_8.rds")

hex_full_count_df_loop2 <- hex_full_count_df_loop2 |>
  dplyr::mutate(type = if_else(std_counts <= 0.20, "low", "high"))

hex_full_count_df_filtered <- hex_full_count_df_loop2 |>
  dplyr::filter(type == "low")

p2_n <-  ggplot(data = hex_full_count_df_loop2, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  #geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) #+
  #annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

p2_n_rm <-  ggplot(data = hex_full_count_df_loop2, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_text(data = hex_full_count_df_filtered, color = "red", aes(x = c_x, y = c_y, group = polygon_id, label = "X", size = 0.5)) +
  #geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) #+
  #annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

cell_count_plot_s_curve <- ggplot(df_bin_centroids_umap_s_curve, aes(x = reorder(as.factor(hexID), -std_counts), y = std_counts)) +
  geom_quasirandom() + xlab("hexagonal id") + ylab("Standardized cell count") +
  geom_hline(yintercept = 0.20, colour = "#de2d26") +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 7),
        axis.text.x = element_text(angle = 90))

df_bin_centroids_umap_s_curve <- df_bin_centroids_umap_s_curve |>
  dplyr::filter(std_counts > 0.20)

## To find the benchmark value
benchmark_umap_s_curve <- find_benchmark_value(.data = distance_umap_s_curve, distance_col = distance)

trimesh_umap_s_curve_rm_low <- ggplot(df_bin_centroids_umap_s_curve, aes(x = x, y = y)) +
  geom_segment(data = tr_from_to_df_umap_s_curve, aes(x = x_from, y = y_from, xend = x_to, yend = y_to)) +
  geom_point(size = 2, colour = "#33a02c") +
  coord_equal()

trimesh_umap_s_curve_rm_low <- trimesh_umap_s_curve_rm_low +
  coord_cartesian(xlim =c(-5, 7.5), ylim = c(-10, 10)) +
  #ggtitle("(a)") +
  xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  theme_light() +
  theme(legend.position = "none", plot.title = element_text(size = 5, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()#change legend key width
  ) #+
  #annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

```{r}
#| eval: false
#| echo: false
#| warning: false
#| fig-cap: Goodness of f
#| label: fig-rmlowdenshex
#| fig-pos: H
##| out-width: 100%

p2_n + p2_n_rm + trimesh_umap_s_curve_rm_low +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```

-->

<!--Addressing low-density hexagons is a systematic process to handle sparsely represented data in certain regions. For each hex bin, we identify the six nearest hex bins using an equal 2D distance metric. Then, we calculate the mean density, as outlined in the equations:

$$
\text{mean density} = \frac{\text{standard count}}{6} 
$${#eq-equationp3}

The standard count is derived from the number of observations in the hex bins. By examining the distribution of mean densities and designating the first quartile as the benchmark value, hex bins with mean densities below this benchmark are removed. This process ensures the elimination of regions with insufficient data density, focusing on areas with more significant data representation and preserving the overall structure in the low-dimensional space.-->

<!--lwd smoothing criteria 1 with s-curve-->
<!--
```{r}
#| warning: false
#| echo: false
#| eval: false

MSE_df_1_lwd <- read_rds("data/s_curve/s_curve_summary_lwd_tsne.rds")
MSE_df_2_lwd <- read_rds("data/s_curve/s_curve_summary_lwd_umap.rds")
MSE_df_3_lwd <- read_rds("data/s_curve/s_curve_summary_lwd_phate.rds")
MSE_df_4_lwd <- read_rds("data/s_curve/s_curve_summary_lwd_trimap.rds")
MSE_df_5_lwd <- read_rds("data/s_curve/s_curve_summary_lwd_pacmap.rds")

MSE_df_lwd <- dplyr::bind_rows(MSE_df_1_lwd, MSE_df_2_lwd, MSE_df_3_lwd, MSE_df_4_lwd, MSE_df_5_lwd)

MSE_df_lwd$method <- factor(MSE_df_lwd$method, levels = c("tSNE", "UMAP", "PHATE", "TriMAP", "PaCMAP"))

## To draw with AIC
aic_plot_lwd <- ggplot(MSE_df_lwd |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_hex,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0.2, linetype="dashed", 
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("benchmark value")
## Effective number of bins along x-axis

mse_plot_lwd <- ggplot(MSE_df_lwd |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_hex,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0.2, linetype="dashed", 
                color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("benchmark value")

```{r}
#| eval: false
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different benchmark values to remove the low-density heaxgons. What is the effective benchmark value to remove the low-density heaxgons? The MSE plot have a steep slope at the beginning, indicating that a smaller benchmark causes a small amount of error. Then, the slope gradually increases or level up, indicating that a higher number of benchmark values generates a higher error. Using the reverse elbow method, when the benchmark value is set to 0.2, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that higher benchmark values remove the necessary bins as well which lead to distruct the 2D structure."
#| label: fig-diagnosticpltScurvelwd
##| out-width: 100%
#| fig-pos: H

aic_plot_lwd + mse_plot_lwd +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```
-->

@@

```{r}
#| echo: false
#| fig-cap: "Distribution of 2D Euclidean distances."
#| label: fig-distScurve

@@

Visualizing the models alongside the original high-D data provides insights into how different techniques capture the underlying clustering structure. The tSNE model exhibits five well-separated clusters, effectively preserving both local and global structures (see video link of @fig-modelfiveGau (a)). UMAP, on the other hand, also presents five distinct clusters but with a more flattened surface appearance (see video link of @fig-modelfiveGau (b)). The PHATE model shows five separated clusters resembling triangles or partial triangles but struggles to capture local structures (see video link of @fig-modelfiveGau (c)). TriMAP and PaCMAP both display five well-separated clusters with flat surfaces, each capturing within-cluster variation to varying extents. Despite the various 2D representations, all NLDR techniques preserve the global structure (see video link of @fig-modelfiveGau (d), (e) respectively). However, tSNE, effectively capture both local and global structures, as indicated by lower AIC values in @fig-diagnosticpltGau (a).

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelfiveGau
#| fig-cap: "Is there a best model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model in the 2D space with tSNE (<https://youtu.be/x5VPB-wOLm4>), (b) Model in the 2D space with UMAP (<https://youtu.be/Cm1dW6iLyHQ>), (c) Model in the 2D space with PHATE (<https://youtu.be/HB0Y3ilz6sI>), (d) Model in the 2D space with TriMAP (<https://youtu.be/iSVkXIrOrSA>), and (e) Model in the 2D space with PaCMAP (<https://youtu.be/O2QBctsc4uM>)."
#| out-width: 100%

@@

<!---

## tSNE

shape_value_gau <- calculate_effective_shape_value(.data = tSNE_data_gau,
                                                   x = tSNE1, y = tSNE2)

num_bins_vec <- 1:20 ## Number of bins along the x-axis

vec <- stats::setNames(rep("", 4), c("number_of_bins", "number_of_observations", "total_error", "total_mse"))  ## Define column names

eval_data_training <- dplyr::bind_rows(vec)[0, ]
eval_data_training <- eval_data_training |>
  dplyr::mutate_if(is.character, as.numeric)

for (i in 1:length(num_bins_vec)) {
  model_object <- fit_high_d_model(training_data = training_data_1, nldr_df_with_id = tSNE_data_gau, x = "tSNE1", y = "tSNE2", num_bins_x = num_bins_vec[i], shape_val = shape_value_gau,
                                 is_bin_centroid = TRUE,
                                 is_rm_lwd_hex = FALSE,
                                 benchmark_to_rm_lwd_hex = NA,
                                 is_avg_high_d = TRUE, column_start_text = "x")

  pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "tSNE")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

  eval_df_training <- eval_df_training |>
    mutate(num_bins_x = num_bins_vec[i])

  eval_data_training <- dplyr::bind_rows(eval_data_training, eval_df_training)
}

## Add new column with data types

eval_data_training <- eval_data_training |>
  mutate(data_type = "training")

MSE_df_1_gau <- eval_data_training |>
  dplyr::mutate(method = "tSNE")
```

## UMAP
## Prediction

shape_value_gau <- calculate_effective_shape_value(.data = UMAP_data_gau,
                                                   x = UMAP1, y = UMAP2)

num_bins_vec <- 1:20 ## Number of bins along the x-axis

vec <- stats::setNames(rep("", 4), c("number_of_bins", "number_of_observations", "total_error", "total_mse"))  ## Define column names

eval_data_training <- dplyr::bind_rows(vec)[0, ]
eval_data_training <- eval_data_training |>
  dplyr::mutate_if(is.character, as.numeric)

for (i in 1:length(num_bins_vec)) {

  model_object <- fit_high_d_model(training_data = training_data_1, nldr_df_with_id = UMAP_data_gau, x = "UMAP1", y = "UMAP2", num_bins_x = num_bins_vec[i], shape_val = shape_value_gau,
                                 is_bin_centroid = TRUE,
                                 is_rm_lwd_hex = FALSE,
                                 benchmark_to_rm_lwd_hex = NA,
                                 is_avg_high_d = TRUE, column_start_text = "x")

  pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "UMAP")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

  eval_df_training <- eval_df_training |>
    mutate(num_bins_x = num_bins_vec[i])

  eval_data_training <- dplyr::bind_rows(eval_data_training, eval_df_training)

## Add new column with data types

eval_data_training <- eval_data_training |>
  mutate(data_type = "training")

MSE_df_2_gau <- eval_data_training |>
  dplyr::mutate(method = "UMAP")
```

```{r}
#| eval: false
#| warning: false
#| echo: false
## PAHTE
## Prediction

shape_value_gau <- calculate_effective_shape_value(.data = PHATE_data_gau,
                                                   x = PHATE1, y = PHATE2)

num_bins_vec <- 1:20 ## Number of bins along the x-axis

vec <- stats::setNames(rep("", 4), c("number_of_bins", "number_of_observations", "total_error", "total_mse"))  ## Define column names

eval_data_training <- dplyr::bind_rows(vec)[0, ]
eval_data_training <- eval_data_training |>
  dplyr::mutate_if(is.character, as.numeric)

for (i in 1:length(num_bins_vec)) {

  model_object <- fit_high_d_model(training_data = training_data_1, nldr_df_with_id = PHATE_data_gau, x = "PHATE1", y = "PHATE2", num_bins_x = num_bins_vec[i], shape_val = shape_value_gau,
                                 is_bin_centroid = TRUE,
                                 is_rm_lwd_hex = FALSE,
                                 benchmark_to_rm_lwd_hex = NA,
                                 is_avg_high_d = TRUE, column_start_text = "x")

  pred_df_training <- predict_2d_embeddings(test_data = training_data_1, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "PHATE")

  eval_df_training <- generate_eval_df(data = df_2, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

  eval_df_training <- eval_df_training |>
    mutate(num_bins_x = num_bins_vec[i])

  eval_data_training <- dplyr::bind_rows(eval_data_training, eval_df_training)

## Add new column with data types

eval_data_training <- eval_data_training |>
  mutate(data_type = "training")

MSE_df_3_gau <- eval_data_training |>
  dplyr::mutate(method = "PHATE")

## TriMAP

## Prediction

shape_value_gau <- calculate_effective_shape_value(.data = TriMAP_data_gau,
                                                   x = TriMAP1, y = TriMAP2)

num_bins_vec <- 1:20 ## Number of bins along the x-axis

vec <- stats::setNames(rep("", 4), c("number_of_bins", "number_of_observations", "total_error", "total_mse"))  ## Define column names

eval_data_training <- dplyr::bind_rows(vec)[0, ]
eval_data_training <- eval_data_training |>
  dplyr::mutate_if(is.character, as.numeric)

for (i in 1:length(num_bins_vec)) {

>>>>

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)

## remove low-density hexagons

@@

hex_full_count_df_loop2 <- hex_grid_with_counts |>
  dplyr::mutate(type = if_else(std_counts <= 0.1, "low", "high"))

hex_full_count_df_filtered <- df_bin_centroids |>
  dplyr::mutate(type = if_else(std_counts <= 0.2, "low", "high")) |>
  dplyr::filter(type == "low")

p2_n <-  ggplot(data = hex_full_count_df_loop2, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  #geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) #+
  #annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

p2_n_rm <-  ggplot(data = hex_full_count_df_loop2, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data = hex_full_count_df_filtered, color = "red", aes(x = c_x, y = c_y, group = hexID, label = "X", size = 0.5)) +
  #geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) #+
  #annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

@@

cell_count_plot <- ggplot(df_bin_centroids_s_curve, aes(x = reorder(as.factor(hexID), -std_counts), y = std_counts)) +
  geom_quasirandom() + xlab("hexagonal id") + ylab("Standardized cell count") +
  geom_hline(yintercept = 0.2, colour = "#de2d26") +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 7),
        axis.text.x = element_text(angle = 90))

benchmark_value_rm_lwd <- stats::quantile(df_bin_centroids_s_curve$std_counts, 
                probs = c(0,0.25,0.5,0.75,1), names = FALSE)[2]

df_bin_centroids_low <- df_bin_centroids_s_curve |>
  dplyr::filter(std_counts <= benchmark_value_rm_lwd)

identify_rm_bins <- find_low_dens_hex(df_bin_centroids_all = df_bin_centroids_s_curve, 
                                      num_bins_x = num_bins_x_s_curve, 
                                      df_bin_centroids_low = df_bin_centroids_low)

df_bin_centroids_s_curve <- df_bin_centroids_s_curve |>
  dplyr::filter(!(hexID %in% identify_rm_bins))

pcentroid_plot_s_curve <- ggplot(data = hex_grid, aes(x = x, y = y)) +
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = df_bin_centroids_s_curve, aes(x = c_x, y = c_y), color = "#33a02c") +
  coord_equal() +
  theme_linedraw() +
  theme(legend.position = "bottom", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

p2_n_rm_correct <-  ggplot(data = hex_full_count_df_loop2, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data = df_bin_centroids_low |> dplyr::filter(hexID %in% identify_rm_bins), color = "red", aes(x = c_x, y = c_y, group = hexID, label = "X"), size = 0.5) +
  #geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) #+
  #annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

@@

# 
# distance_plot

@@

distance_plot <- plot_dist(distance_s_curve) +
   geom_hline(yintercept = benchmark_s_curve, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12))

@@

trimesh_removed_s_curve_umap <- trimesh_removed_s_curve_umap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

# df_all_s_curve <- dplyr::bind_cols(training_data |> dplyr::select(-ID), umap_data_with_hb_id)
# 
# show_langevitour(df_all_s_curve, df_bin_s_curve, df_bin_centroids_s_curve, benchmark_s_curve, distance_s_curve, "distance", col_start = "x")
```

@@

```{r}
#| echo: false
#| fig-cap: "Distribution of 2D Euclidean distances."
#| label: fig-stdctsScurve
#| fig-pos: H

@@

```{r}
#| warning: false
#| echo: false
#| message: false
num_bins_x_s_curve <- 6
num_bins_y_s_curve <- 14
hex_size_s_curve <- 0.13

vec <- stats::setNames(rep("", 2), c("benchmark_val_lwd", "mse"))  ## Define column names

centroid_df_training_all <- model_object$df_bin_centroids
avg_df_training_all <- model_object$df_bin

benchmark_rm_hex_vec <- seq(0, 1, by=0.1)
benchmark_rm_hex_vec <- append(benchmark_rm_hex_vec[1:10],0.99)

for (i in 1:length(benchmark_rm_hex_vec)) {

  centroid_df_training <- centroid_df_training_all |>
    dplyr::filter(std_counts > benchmark_rm_hex_vec[i])

  avg_df_training <- avg_df_training_all |>
    dplyr::filter(hb_id %in% centroid_df_training$hexID)

  mse_df <- mse_df |>
    tibble::add_row(benchmark_val_lwd = benchmark_rm_hex_vec[i],
                    mse = eval_list$mse)

mse_plot_lwd <- ggplot(mse_df, aes(x = benchmark_val_lwd,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 0.2, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("benchmark value to remove low-density hexagons")

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different benchmark values to remove the low-density heaxgons. What is the effective benchmark value to remove the low-density heaxgons? The MSE plot have a steep slope at the beginning, indicating that a smaller benchmark causes a small amount of error. Then, the slope gradually increases or level up, indicating that a higher number of benchmark values generates a higher error. Using the reverse elbow method, when the benchmark value is set to 0.2, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that higher benchmark values remove the necessary bins as well which lead to distruct the 2D structure."
#| label: fig-diagnosticpltScurvelwd
##| out-width: 100%
#| fig-pos: H

mse_plot_lwd
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: Goodness of f
#| label: fig-rmlowdenshex
#| fig-pos: H
##| out-width: 100%

p2_n + p2_n_rm + p2_n_rm_correct +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

```{r}
#| echo: false
#| fig-cap: "Distribution of 2D Euclidean distances."
#| label: fig-distScurve
#| fig-pos: H

@@

Visualizing the models alongside the original high-D data provides insights into how different techniques capture the underlying clustering structure. The tSNE model exhibits five well-separated clusters, effectively preserving both local and global structures (see video link of @fig-modelfiveGau (a)). UMAP, on the other hand, also presents five distinct clusters but with a more flattened surface appearance (see video link of @fig-modelfiveGau (b)). The PHATE model shows five separated clusters resembling triangles or partial triangles but struggles to capture local structures (see video link of @fig-modelfiveGau (c)). TriMAP and PaCMAP both display five well-separated clusters with flat surfaces, each capturing within-cluster variation to varying extents. Despite the various 2D representations, all NLDR techniques preserve the global structure (see video link of @fig-modelfiveGau (d), (e) respectively). However, tSNE, effectively capture both local and global structures.

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelfiveGau
#| fig-cap: "Is there a best model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model generated with 21 and 28 bins along the x and y axes, respectively, using a hexagonal size of 0.03 in the 2D space with tSNE, (b) Model generated with 60 and 79 bins along the x and y axes, respectively, using a hexagonal size of 0.01 in the 2D space with UMAP, (c) Model generated with 60 and 1927 bins along the x and y axes, respectively, using a hexagonal size of 0.01 in the 2D space with PHATE, (d) Model generated with 46 and 61 bins along the x and y axes, respectively, using a hexagonal size of 0.013 in the 2D space with TriMAP, and (e) Model generated with 31 and 40 bins along the x and y axes, respectively, using a hexagonal size of 0.02 in the 2D space with PaCMAP."
#| out-width: 100%

@@

::: {#fig-gau1_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_tsne_1.png){#fig-gau1_sc1}

![](figures/five_gau_clusters/sc_tsne_2.png){#fig-gau1_sc2}

![](figures/five_gau_clusters/sc_tsne_3.png){#fig-gau1_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with tSNE in high-D, a video of the tour animation is available at (<https://youtu.be/oQxEb4wRdHI>).
:::

::: {#fig-gau2_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_umap_1.png){#fig-gau2_sc1}

![](figures/five_gau_clusters/sc_umap_2.png){#fig-gau2_sc2}

![](figures/five_gau_clusters/sc_umap_3.png){#fig-gau2_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with UMAP in high-D, a video of the tour animation is available at (<https://youtu.be/JW49csPpDx4>).
:::

::: {#fig-gau3_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_phate_1.png){#fig-gau3_sc1}

![](figures/five_gau_clusters/sc_phate_2.png){#fig-gau3_sc2}

![](figures/five_gau_clusters/sc_phate_3.png){#fig-gau3_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with PHATE in high-D, a video of the tour animation is available at (<https://youtu.be/xqhRODt8Z-o>).
:::

::: {#fig-gau4_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_trimap_1.png){#fig-gau4_sc1}

![](figures/five_gau_clusters/sc_trimap_2.png){#fig-gau4_sc2}

![](figures/five_gau_clusters/sc_trimap_3.png){#fig-gau4_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with TriMAP in high-D, a video of the tour animation is available at (<https://youtu.be/X-uHSB4VoZ0>).
:::

::: {#fig-gau5_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_pacmap_1.png){#fig-gau5_sc1}

![](figures/five_gau_clusters/sc_pacmap_2.png){#fig-gau5_sc2}

![](figures/five_gau_clusters/sc_pacmap_3.png){#fig-gau5_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with PaCMAP in high-D, a video of the tour animation is available at (<https://youtu.be/UxvU8HQd7JM>).
:::
